---
title: "Basic data manipulation with mzDataTable"
author: "Patrick Brophy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mzDataTable-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The [mzDataTable](https://github.com/pmbrophy/mzDataTable) package aims to provide methods get .mzML/.mzXML data into [data.table](https://cran.r-project.org/web/packages/data.table/) format through the use of [mzR](https://www.bioconductor.org/packages/release/bioc/html/mzR.html). There are three conversion functions provided:

1. `mzML2dataTable` reads an .mzML/.mzXML file into memory as a `data.table`
2. `mzML2csv` converts an .mzML/.mzXML file to a .csv file by first reading chunks of the .mzML/.mzXML file into memory as `data.table`s and sequentially writing them to the .csv file.
3. `mzML2diskFrame` converts an .mzML/.mzXML file to a `disk.frame` directory by first reading chunks of the .mzML/.mzXML file into memory as `data.table`s and sequentially writing them as `disk.frame` chunks. This provides efficent on-disk manipulation of raw mass spectrometry data.

# Import and Conversion Functions
## The mzML2dataTable Function: get data into memory

We can use the mock .mzML data provided in the [mzDataTable](https://github.com/pmbrophy/mzDataTable) package as an example or get real data from [msdata](https://bioconductor.org/packages/msdata/). 
```{r setup}
#load the library and data.table
library(mzDataTable)
library(data.table)

dataPath <- msdata::proteomics(full.names = TRUE)[3]

#Read the .mzML file into memory as a data.table
dt <- mzML2dataTable(path = dataPath, scans = c(1:10))

is.data.table(dt)

head(dt)
```
