---
title: "Basic data manipulation with mzDataTable"
author: "Patrick Brophy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mzDataTable-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
#load the library and data.table
library(mzDataTable)
```

# Introduction

The [mzDataTable](https://github.com/pmbrophy/mzDataTable) package aims to provide methods get .mzML/.mzXML data into [data.table](https://cran.r-project.org/web/packages/data.table/) format through the use of [mzR](https://www.bioconductor.org/packages/release/bioc/html/mzR.html). There are three conversion functions provided:

1. `mzML2dataTable` reads an .mzML/.mzXML file into memory as a `data.table`
2. `mzML2csv` converts an .mzML/.mzXML file to a .csv file by first reading chunks of the .mzML/.mzXML file into memory as `data.table`s and sequentially writing them to the .csv file.
3. `mzML2diskFrame` converts an .mzML/.mzXML file to a `disk.frame` directory by first reading chunks of the .mzML/.mzXML file into memory as `data.table`s and sequentially writing them as `disk.frame` chunks. This provides efficent on-disk manipulation of raw mass spectrometry data.

# Import and Conversion Functions
## The `mzML2dataTable` Function: get data into memory

We can use the mock .mzML data provided in the [mzDataTable](https://github.com/pmbrophy/mzDataTable) package as an example or get real data from [msdata](https://bioconductor.org/packages/msdata/). 
```{r}
dataPath <- msdata::proteomics(full.names = TRUE)[3]

#Read the .mzML file into memory as a data.table
dt <- mzML2dataTable(path = dataPath)

data.table::is.data.table(dt)

head(dt)
```

## The `mzML2diskFrame` Function: get serialized data onto hard disk for on-disk manipulations
```{r}
dskF_path <- paste0(getwd(), "/", basename(dataPath),".df")

#Create disk frame
#dskF <- mzML2diskFrame(path = dataPath, diskFramePath = dskF_path)

#Connect to disk frame
dskF <- disk.frame::disk.frame(path = dskF_path)
disk.frame::setup_disk.frame()
options(future.globals.maxSize = Inf)
```

# Data Manipulation Functions

Once data are imported into R, there are some functions provided for basic manipulations:

1. `getTIC` calculates the total ion chromatogram
2. `getBPI` calculates the base peak intensity chromatogram
3. `getXIC` extracts summed ion signal intensity over a specified mass range 
4. `getSumSpectrum_dt` produces a summed (and optionally normalized) mass spectrum

## `getTIC` example: plot total ion chromatogram from MS level == 1, 2, and 3 individually
```{r}
tic_dt <- getTIC(mzObj = dt[msLevel == 1], normalize = FALSE)
tic_df <- getTIC(mzObj = dskF[msLevel == 1], normalize = FALSE)

identical(tic_dt, tic_df)

tic_dt <- getTIC(mzObj = dt[msLevel == 1], normalize = TRUE)
tic_df <- getTIC(mzObj = dskF[msLevel == 1], normalize = TRUE)

identical(tic_dt, tic_df)

tic_dt <- getTIC(mzObj = dt, normalize = TRUE)
tic_df <- getTIC(mzObj = dskF, normalize = TRUE)

identical(tic_dt, tic_df)

plot(x = tic_dt$retentionTime, y = tic_dt$intensity, type = "l")
```

## `getBPI` example: get base peak intensity
```{r}
bpi_dt <- getBPI(mzObj = dt[msLevel == 1], normalize = FALSE)
bpi_df <- getBPI(mzObj = dskF[msLevel == 1], normalize = FALSE)

identical(bpi_dt, bpi_df)

bpi_dt <- getBPI(mzObj = dt[msLevel == 1], normalize = TRUE)
bpi_df <- getBPI(mzObj = dskF[msLevel == 1], normalize = TRUE)

identical(bpi_dt, bpi_df)

remove(bpi_dt, bpi_df)
bpi_dt <- getBPI(mzObj = dt, normalize = TRUE)
bpi_df <- getBPI(mzObj = dskF, normalize = TRUE)

identical(bpi_dt, bpi_df)

plot(x = bpi_dt$retentionTime, y = bpi_dt$intensity, type = "l")
```

## `getXIC` example: extract m/z 376.2756 +/- 25 ppm
```{r}

xic_dt <- getXIC(mzObj = dt[msLevel == 1], mz = 376.2756, ppmTol = 25, iStart = 1)

xic_dt <- getXIC(mzObj = dt[msLevel == 1], mz = 376.2756, ppmTol = 25)
xic_df <- getXIC(mzObj = dskF[msLevel == 1], mz = 376.2756, ppmTol = 25)

identical(xic_dt, xic_df)

xic_dt <- getXIC(mzObj = dt, mz = 376.2756, ppmTol = 25)
xic_df <- getXIC(mzObj = dskF, mz = 376.2756, ppmTol = 25)

identical(xic_dt, xic_df)

xic_dt <- getXIC(mzObj = dt, mz = 376.2756, ppmTol = 25, iStart = 10, iStop = 12)
xic_df <- getXIC(mzObj = dskF, mz = 376.2756, ppmTol = 25, iStart = 10, iStop = 12)

identical(xic_dt, xic_df)

xic_dt <- getXIC(mzObj = dt, mz = 376.2756, ppmTol = 25, tStart = 2730, tStop = 2750)
xic_df <- getXIC(mzObj = dskF, mz = 376.2756, ppmTol = 25, tStart = 2730, tStop = 2750)

identical(xic_dt, xic_df)

plot(x = xic_dt$retentionTime, y = xic_dt$intensity, type = "l")
```

## `getSumSpectrum_dt` example: sum MS1 spectra
```{r}
spec_norm <- getSumSpectrum(mzObj = dt[msLevel == 1], normalize = TRUE, isCentroid = FALSE)
spec <- getSumSpectrum(mzObj = dt[msLevel == 1], normalize = FALSE, isCentroid = FALSE)

spec_dt <- getSumSpectrum(mzObj = dt, normalize = FALSE, isCentroid = TRUE, ppmTol = 100)
spec_dskF <- getSumSpectrum(mzObj = dskF, normalize = FALSE, isCentroid = TRUE, ppmTol = 100)
identical(spec_dt, spec_dskF)

spec_dt <- getSumSpectrum(mzObj = dt, normalize = FALSE, isCentroid = TRUE, ppmTol = 100, iStart = 10, iStop = 20)
spec_dskF <- getSumSpectrum(mzObj = dskF, normalize = FALSE, isCentroid = TRUE, ppmTol = 100,  iStart = 10, iStop = 20)


spec <- getSumSpectrum(mzObj = dt, normalize = FALSE, isCentroid = TRUE, ppmTol = 100, tStart = 10, tStop = 20)
spec <- getSumSpectrum(mzObj = dskF, normalize = FALSE, isCentroid = TRUE, ppmTol = 100,  tStart = 10, tStop = 20)

plot(x = spec_dskF$mzGrid, y = spec_dskF$intensity, type = "h")
```

###UNDERSTAND disk.frame OPERATION ORDER
```{r eval=FALSE, include=FALSE}
#What happens? 1) read mz, seqNum, intensity from disk as data.table. 2) apply %between%
dskF_test <- dskF[seqNum %between% c(1,2), keep = c("mz", "seqNum", "intensity")]

#Is it the same as this?
dskF_test <- dskF[keep = c("mz", "seqNum", "intensity")]

dskF_test <- dskF_test[seqNum %between% c(1,2)]

#What happens with 1 vs 6 workers? 
#1) each worker reads in mz, seqNum, intensity
#2) each worker then subsets by %between%?

#So, in the case of a single worker... read in columns, subset, return subset to Global Env, and repeat?
```
